---
title: "Discover and inspect data packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Discover and inspect data packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
live <- identical(Sys.getenv("IN_PKGDOWN"), "true") &&
  !identical(Sys.getenv("GLCDP_SKIP_LIVE"), "true")
```

This article shows how to choose a Global Light Commons package and understand
its contents before downloading or importing measurements. Remote examples run
on the pkgdown website but remain unevaluated in ordinary package and CRAN
builds. Set `GLCDP_SKIP_LIVE=true` to request an offline website build.

```{r load}
library(glcdp)
```

## Understand the registry

`glc_packages()` returns one row per registered repository. The most useful
fields distinguish the repository's current revision from its most recent
passing revision:

- `current_status` and `current_commit` describe the configured current
  revision;
- `latest_pass_commit` and `has_latest_pass` identify the last validated
  revision available for reproducible use; and
- `attestation_verified` reports whether the registry attestation was
  verified.

```{r registry, eval = live}
packages <- glc_packages()
packages[, c(
  "id", "repository", "current_status", "has_latest_pass",
  "attestation_verified"
)]
```

Registry results are cached for the R session. Set `refresh = TRUE` only when
you need to fetch the registry again.

```{r registry-refresh, eval = live}
packages <- glc_packages(refresh = TRUE)
```

Searches are fixed and case-insensitive by default:

```{r registry-search, eval = live}
glc_search_packages("guidolin", packages)
glc_search_packages(packages = packages, status = c("pass", "fail"))
glc_search_packages(packages = packages, has_pass = FALSE)
```

## Choose a revision deliberately

Opening a registered package with the default `ref = "latest_pass"` selects an
exact passing commit, not a moving branch:

```{r open-latest-pass, eval = live}
x <- glc_open("tscnlab/guidolin-glee-datasetv2")
x
```

Use `ref = "current"` when you explicitly need the registry's current
revision. If that revision is not passing, `glcdp` warns. You can also provide
an exact 40-character commit SHA; commits that are not selected through a
registry record are marked as unverified.

```{r open-revisions, eval = live}
current <- glc_open(
  "tscnlab/guidolin-glee-datasetv2",
  ref = "current"
)
current

registry_row <- glc_search_packages("guidolin", packages)
registry_row$repository[[1]]
registry_row$latest_pass_commit[[1]]

pinned <- glc_open(
  registry_row$repository[[1]],
  ref = registry_row$latest_pass_commit[[1]]
)
pinned
```

For private repositories, pass `token` directly or define `GITHUB_PAT` or
`GITHUB_TOKEN`. Do not put tokens in scripts, vignettes, or package options.

## Start with a package summary

`glc_summary()` reports the schema version and counts of studies, datasets,
participants, devices, file groups, files, and variables. It also summarizes
modalities, time zones, and primary variables.

```{r summary, eval = live}
glc_summary(x)
```

## Explore inventories

The inventories are tibbles, so they can be printed, filtered, or joined using
ordinary data-frame tools.

```{r resources, eval = live}
glc_resources(x)
```

`glc_datasets()` describes logical datasets and their associations. Once you
have a dataset id, reuse it to narrow the other inventories.

```{r datasets, eval = live}
glc_datasets(x)

glc_files(x, dataset_id = "DS001")
glc_files(
  x,
  dataset_id = "DS001",
  modality = "light",
  available = TRUE
)
```

File inventories expose both declared and resolved paths, format, encoding,
time zone, role, data state, device, storage type, expected size, and
availability. Git LFS-backed files are identified without requiring a local
Git LFS installation.

Variable inventories can be narrowed by dataset, file group, semantic term, or
primary status:

```{r variables, eval = live}
glc_variables(x, dataset_id = "DS001")
glc_variables(x, dataset_id = "DS001", primary = TRUE)
glc_variables(x, term = "photopic illuminance")
```

Use source variable names with the `variables` argument of `glc_read()` and
semantic terms with its `terms` argument.

## Load and search metadata

With no `resources` argument, `glc_metadata()` loads the core resources that
the package declares. Requesting resources explicitly is often faster and
makes dependencies clearer.

```{r metadata, eval = live}
metadata <- glc_metadata(x, resources = c("study", "participants"))
names(metadata)
metadata$study
metadata$participants
```

JSON objects remain lists, tabular resources become tibbles, and directory
resources become named lists keyed by package-relative path.

Search traverses nested metadata and reports the resource, record, complete
field path, value, and context for each match:

```{r metadata-search, eval = live}
glc_search_metadata(x, "light exposure")

glc_search_metadata(
  x,
  "age|chronotype",
  resources = "participant_characteristics",
  fixed = FALSE
)

city_fields <- glc_search_metadata(
  x,
  "city",
  resources = "study",
  search_in = "fields"
)
unique(city_fields$field)

glc_search_metadata(
  x,
  "Munich",
  resources = "study",
  fields = "contributor_institution_city"
)
```

## Work with local packages

Local packages use the same public interface, which makes them useful for
development, validation follow-up, and offline analysis:

```{r local, eval = FALSE}
x_local <- glc_open("path/to/package", quiet = TRUE)
glc_summary(x_local)
glc_files(x_local, available = FALSE)
```

The directory must contain a `datapackage.json` descriptor and all paths needed
by the selected operation. A subset created by `glc_download()` can be reopened
in exactly the same way.

After selecting and collecting compatible light data, use the
[LightLogR function reference](https://tscnlab.github.io/LightLogR/reference/index.html)
for downstream quality checks, summaries, metrics, and visualizations.
