---
title: "Get started with glcdp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with glcdp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
live <- identical(Sys.getenv("IN_PKGDOWN"), "true") &&
  !identical(Sys.getenv("GLCDP_SKIP_LIVE"), "true")
```

`glcdp` provides a focused route from a Global Light Commons (GLC) package to
analysis-ready R data:

1. discover a registered package;
2. open an immutable revision;
3. inspect its datasets, files, variables, and metadata;
4. read only the data you need; and
5. collect compatible file groups for analysis.

Remote examples run when pkgdown builds the package website. They are displayed
without execution during ordinary package and CRAN builds, which keeps those
checks independent of network availability. Website maintainers can also set
`GLCDP_SKIP_LIVE=true` for an explicitly offline pkgdown build.

## Install and load

Install the development version from GitHub and attach the package:

```{r install, eval = FALSE}
pak::pak("tscnlab/glc-dp-r")
library(glcdp)
```

`glcdp` currently understands the following GLC schemas:

```{r schemas}
library(glcdp)
glc_schema_versions()
```

## Discover a package

The registry includes both passing and non-passing current revisions. Keeping
both visible makes validation status explicit instead of silently hiding
packages with problems.

```{r discover, eval = live}
packages <- glc_packages()
packages

glc_search_packages("guidolin", packages)
```

You can also filter on validation status or on whether a package has a recorded
passing revision:

```{r filter-registry, eval = live}
glc_search_packages(packages = packages, status = "pass")
glc_search_packages(packages = packages, has_pass = TRUE)
```

## Open a reproducible revision

Registered packages open at their latest passing commit by default. The
returned handle records the repository, exact commit, schema version, and
whether the revision was verified against the registry.

```{r open, eval = live}
guidolin <- glc_open("tscnlab/guidolin-glee-datasetv2")
guidolin
```

The same function opens a local package directory or its `datapackage.json`
file:

```{r open-local, eval = FALSE}
local_package <- glc_open("path/to/data-package")
```

## Inspect before reading

A compact summary is a useful first look:

```{r inspect-summary, eval = live}
glc_summary(guidolin)
```

The inventories make data selection explicit. List datasets, then narrow the
file and variable inventories to the dataset you intend to read.

```{r inspect-inventories, eval = live}
glc_datasets(guidolin)
glc_files(guidolin, dataset_id = "DS001")
glc_variables(guidolin, dataset_id = "DS001", primary = TRUE)
```

Use `glc_metadata()` for structured metadata and `glc_search_metadata()` when
you need to locate a value without knowing its resource or field in advance.

```{r inspect-metadata, eval = live}
metadata <- glc_metadata(guidolin, resources = c("study", "participants"))
metadata$study
metadata$participants

glc_search_metadata(guidolin, "light", resources = "study")
glc_search_metadata(guidolin, "light", resources = "datasets")
```

## Read selected data

`glc_read()` follows the package metadata when it chooses headers, column
types, factor levels, datetime formats, decimal marks, encodings, and time
zones. A dataset selection is required so that a large package is not imported
accidentally.

```{r read, eval = live}
collection <- glc_read(
  guidolin,
  dataset_id = "DS001",
  primary_only = TRUE
)
collection
```

In interactive sessions, `glc_read()` displays progress across the selected
files. Set `progress = FALSE` to suppress the indicator, for example in a
script that manages its own progress reporting.

The result has one row per compatible file group and stores each imported table
in its `data` list-column. Inspect or process groups separately when their
roles, modalities, or schemas differ.

```{r inspect-collection, eval = live}
collection$data[[1]]
```

For quick exploration, limit the number of rows read from each file:

```{r sample, eval = live}
sample <- glc_read(
  guidolin,
  dataset_id = "DS001",
  variables = "LIGHT",
  n_max = 100
)
```

## Collect compatible groups

`glc_collect()` checks that the selected groups have compatible columns, types,
time zones, modalities, roles, data states, and datetime specifications before
combining them. Its default output maps the dataset id to `Id`, the participant
id to `participant_Id`, and adds `Datetime` and `file.name`. Internal `.glc_*`
provenance columns are removed from this analysis-ready result, matching the
core conventions described in
[LightLogR's import documentation](https://tscnlab.github.io/LightLogR/reference/import_Dataset.html).

```{r collect, eval = live}
light_data <- glc_collect(collection)
light_data
```

The result can be passed directly to LightLogR. Continue with its guides to
[visualizing light logger data](https://tscnlab.github.io/LightLogR/articles/Visualizations.html)
or [calculating light exposure metrics](https://tscnlab.github.io/LightLogR/articles/Metrics.html).

Use `standardize = "none"` to leave the source columns and `.glc_*` provenance
columns unchanged:

```{r collect-unstandardized, eval = live}
source_data <- glc_collect(sample, standardize = "none")
source_data
```

## Where to go next

- [Discover and inspect data packages](discover-and-inspect.html) covers
  registry, revision, inventory, and metadata workflows.
- [Import and download data](import-and-download.html) covers precise import
  selections, persistent downloads, and reproducibility manifests.
