---
title: "Import and download data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Import and download data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
live <- identical(Sys.getenv("IN_PKGDOWN"), "true") &&
  !identical(Sys.getenv("GLCDP_SKIP_LIVE"), "true")
```

This article covers two related workflows:

- import selected files directly into R with `glc_read()`; and
- create a persistent, reproducible local subset with `glc_download()`.

Remote examples run on the pkgdown website but remain unevaluated in ordinary
package and CRAN builds. Set `GLCDP_SKIP_LIVE=true` to request an offline
website build.

```{r load}
library(glcdp)
```

```{r open, eval = live}
x <- glc_open("tscnlab/guidolin-glee-datasetv2")
```

## Select before importing

Inspect datasets, file groups, and variables first. This avoids transferring
unneeded data and gives you the stable ids and source names used by the read
selectors.

```{r inspect, eval = live}
glc_datasets(x)
glc_files(x, dataset_id = "DS001")
glc_variables(x, dataset_id = "DS001")
```

`dataset_id` is required by `glc_read()`. Use `dataset_id = "all"` only when
you deliberately want every dataset.

```{r read-dataset, eval = live}
collection <- glc_read(x, dataset_id = "DS001")
collection
```

Interactive calls show progress across the selected files by default. Use
`progress = FALSE` to suppress the indicator when a script or application
provides its own progress reporting.

Narrow the import with any combination of file-group ids or indices, file
paths or basenames, source variable names, and semantic terms:

```{r read-selection, eval = live}
file_inventory <- glc_files(x, dataset_id = "DS001")
first_file <- file_inventory[1, ]
first_file
file_group <- first_file$file_group_id[[1]]
file_group
files <- basename(first_file$path[[1]])
files

selected <- glc_read(
  x,
  dataset_id = "DS001",
  file_group = file_group,
  files = files,
  variables = c("LIGHT"),
  n_max = 1000
)
selected$data
```

When `variables` or `terms` are selected, the imported tables contain only the
matching source variables. `glcdp` still uses any required date or time columns
internally to construct `.glc_datetime`, but does not retain them unless the
filters select them. Set `primary_only = TRUE` to select all declared primary
variables.

```{r read-primary, eval = live}
primary <- glc_read(
  x,
  dataset_id = "DS001",
  primary_only = TRUE
)
```

## Understand the imported collection

A `glc_data_collection` has one row per file group. Descriptive columns record
the dataset, participant, device, modality, role, data state, time zone,
datetime specification, and source files. The imported table is in the `data`
list-column.

```{r collection, eval = live}
collection[, setdiff(names(collection), "data")]
collection$data[[1]]
```

Each imported table also contains `.glc_*` provenance columns such as the
dataset id, participant id, source file, and constructed datetime. These make
row origins explicit after tables are combined.

By default, metadata mismatches such as undeclared extra columns or values that
cannot be parsed to the declared type are errors. During exploratory work, use
`problems = "warn"` to retain problematic data with warnings; inspect the
result before analysis.

```{r read-warn, eval = live}
exploratory <- glc_read(
  x,
  dataset_id = "DS001",
  problems = "warn"
)
```

## Collect analysis-ready data

Combine file groups only when their structures and meanings are compatible:

```{r collect, eval = live}
light_data <- glc_collect(collection)
light_data
```

## Extract or add metadata

Use `extract_metadata()` when you want one concise row per imported file group,
and `add_metadata()` when the same fields should be available on every
observation. Both functions require an explicit metadata source and an explicit
field selection. The default key is `file_group_id`, which `glc_collect()` adds
to standardized data. If the input is grouped, the extracted tibble retains
that grouping and includes its grouping columns before `file_group_id`.

```{r metadata-pair, eval = live}
analysis_metadata <- tibble::tibble(
  file_group_id = unique(as.character(light_data$file_group_id)),
  analysis_set = "primary"
)

metadata_summary <- extract_metadata(
  light_data,
  analysis_metadata,
  fields = "analysis_set"
)
metadata_summary

enriched_data <- add_metadata(
  light_data,
  analysis_metadata,
  fields = "analysis_set"
)

enriched_data |> head() |> dplyr::select(-file.name)
```

The metadata source can also be a local CSV/TSV path or the package handle.
For a package handle, `glcdp` follows each file group to its dataset and device,
and follows the dataset to its participant and study. It can therefore assemble
fields such as `participant_age` without manually naming the participant
resource. Dataset-, participant-, and study-level values repeat across file
groups. Use `by = "Id"` when you explicitly want one row per dataset instead.
Field and relationship resolution must remain unambiguous; use `resource` to
restrict field discovery when the same field occurs in multiple connected
resources.

```{r metadata-resource}
# Searches metadata:
glc_search_metadata(x, "age", search_in = "fields")

# Extracts relevant metadata for collected dataset:
dataset_metadata <- extract_metadata(
  light_data,
  x,
  fields = c(
    "dataset_timezone",
    "dataset_location",
    "participant_age",
    "study_title",
    "device_model"
  )
)
dataset_metadata

# Adds relevant metadata into collected dataset
add_metadata(
  light_data,
  x,
  fields = c(
    "dataset_timezone",
    "dataset_location",
    "participant_age",
    "study_title",
    "device_model"
  )
) |>
  head() |>
  dplyr::select(-file.name)
```

When adding project-specific metadata to a data package, store it under a
stable package-relative path such as `data/metadata.csv` and declare it as a
resource in `datapackage.json`. The functions never guess from the working
directory or neighboring files. They error when no identifiers or fields
match, and warn while retaining useful results for partial matches.

The default `standardize = "lightlogr"` follows the data conventions used by
[LightLogR](https://tscnlab.github.io/LightLogR/) and adds:

- `Id`, derived from the dataset id;
- `file_group_id`, the stable dataset file-group id used by the metadata
  helpers;
- `participant_Id`, derived from the participant id;
- `Datetime`, constructed from the metadata-defined datetime specification;
  and
- `file.name`, derived from the source file path.

The standardized result contains no internal `.glc_*` provenance columns. It
is ordered by `Id` and `Datetime` and grouped by `Id`. Use `standardize =
"none"` if you want an ungrouped tibble whose source and provenance columns
are left unchanged.

```{r collect-none, eval = live}
source_data <- glc_collect(collection, standardize = "none")
```

`glc_collect()` refuses to combine groups that differ in columns, types, time
zones, modalities, roles, data states, or datetime specifications. It also
rejects contradictory file-group relationships and multiple device links
within one dataset. Keep those groups separate or select a compatible subset
with `glc_read()`.

Collected data can be used directly with LightLogR's
[data-quality and insight functions](https://tscnlab.github.io/LightLogR/reference/index.html#insight),
[visualization guide](https://tscnlab.github.io/LightLogR/articles/Visualizations.html),
and [metrics guide](https://tscnlab.github.io/LightLogR/articles/Metrics.html).

## Download a reproducible subset

`glc_download()` creates a persistent directory while preserving
package-relative paths. Its safe default downloads only the descriptor, core
metadata, and required schemas:

For public packages, ordinary Git files are transferred from immutable raw
URLs at the selected commit, avoiding per-file GitHub API requests. A token
supplied to `glc_open()` continues to use the authenticated API transport,
including for private repositories.

```{r download-metadata, eval = live}
metadata_dir <- file.path(tempdir(), "guidolin-metadata")
glc_download(x, metadata_dir)
```

Request data explicitly and apply the same kinds of selectors used during
inspection:

```{r download-data, eval = live}
data_dir <- file.path(tempdir(), "guidolin-ds001")
downloads <- glc_download(
  x,
  data_dir,
  include = "data",
  dataset_id = "DS001"
)
downloads
```

Use `include = "all"` only when you intend to mirror every declared resource.
The `resources` and `files` arguments can further narrow a download. Existing
files are protected unless `overwrite = TRUE` is set explicitly.

Every download writes `glcdp-manifest.json`, recording the source repository,
exact commit, registry verification state, schema version, selection, hashes,
storage types, sizes, and Git LFS object ids. Reopen the directory to use the
same inspection and import API without fetching the package again:

```{r reopen, eval = live}
local <- glc_open(data_dir)
glc_summary(local)
glc_files(local, available = FALSE)

local_collection <- glc_read(local, dataset_id = "all")
local_data <- glc_collect(local_collection)
```

The descriptor and core metadata retain the records declared by the source
package. For a local subset, `glc_summary()` distinguishes locally available
datasets, file groups, and files from those declared records. When the package
is incomplete, `glc_read()` also reports how many declared datasets and files
are locally available, then skips absent files. Thus, `dataset_id = "all"`
reads all data included in the subset. Use
`glc_files(local, available = FALSE)` to inspect the omitted file records.

## Temporary reads versus persistent storage

Remote `glc_read()` calls use session-temporary storage by default. This is a
good fit for one-off analysis and leaves no persistent files behind.

Pass `cache_dir` to `glc_open()` when you want remote files reused across
calls, or use `glc_download()` when you want an explicit, portable package
subset with a manifest:

```{r cache, eval = live}
cached <- glc_open(
  "tscnlab/guidolin-glee-datasetv2",
  cache_dir = file.path(tempdir(), "glcdp-cache")
)
```

Choose the cache for performance; choose a downloaded subset for a durable,
inspectable analysis input.
